\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2024-02-05 doc/paper-jss/sdmTMB-paper.Rnw
Edits
Eric/Lewis edits

Data are often collected in space or in space repeatedly over time. While such data are a rich source of information across many fields,
they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables \citep{cressie1993, diggle2007, cressie2011}.
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference \citep{legendre1989a, dormann2007}, can improve predictions \citep[e.g.,][]{shelton2014}, and can be of interest itself \citep[e.g.,][]{thorson2019d, barnett2021}.

@@

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \pkg{fmesher} or \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
\Rev{B2}Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and approximated in discrete time-intervals.

This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \sdmTMB\ is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \sdmTMB\ to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of links to other packages and future development directions (Section~\ref{discussion}).

@@

<<matern-range, warning=FALSE, message=FALSE, fig.width=9, out.width="\\textwidth", fig.asp=0.33, fig.align='center', fig.cap="Example Gaussian random fields for two range values. The range describes the distance at which spatial correlation decays to $\\approx 0.13$ in coordinate units (i.e., the distance at which two points are effectively independent). Panel a shows a shorter range than panel b, which results in a ``wigglier'' surface. Panel c shows the Mat\\'ern function for these two range values. The dashed horizontal line shows a correlation of 0.13.">>=
predictor_dat <- expand.grid(
  x = seq(0, 1, length.out = 100),
  y = seq(0, 1, length.out = 100),
  year = seq_len(6)
)
sim_mesh <- make_mesh(predictor_dat, xy_cols = c("x", "y"), cutoff = 0.01)
s1 <- sdmTMB_simulate(
  formula = ~1,
  data = predictor_dat,
  mesh = sim_mesh,
  range = 0.2,
  phi = 0.1,
  sigma_O = 0.2,
  seed = 1,
  B = 0
)
sim_g1 <- ggplot(s1, aes(x, y, fill = mu)) +
  geom_raster(show.legend = FALSE) +
  scale_fill_viridis_c(option = "C") +
  coord_fixed(expand = FALSE) +
  theme_light() +
  # theme(axis.title = element_blank()) +
  ggtitle("(a) Range = 0.2") +
  labs(x = "X", y = "Y")

@@

Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}).
If we define $\lVert h \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\sigma^2$ is the marginal variance (magnitude of the random field ``wiggles''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
The parameter $\nu$ controls the smoothness of the covariance function.
In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}.
A more interpretable derived parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\mathrm{range} = \sqrt{8} / \kappa$), which has the empirically derived property of the distance at which correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).

@@

The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
\[
\Phi\left( s_j, s_k \right) =
  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \bm{H} \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \bm{H} \right).
\]
Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{haskard2007, lindgren2011, thorson2015ices}.

@@

\Rev{B4b}For a user of \sdmTMB, the following are the important elements to understand.
First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
This means that GRF models can be computationally approximated as GMRFs.
By working with GMRFs, one can take advantage of theory developed to estimate their sparse precision matrix efficiently \citep{rue2005gmrf, lindgren2011} and avoid the inversion of large dense matrices.
Second, the SPDE involves piece-wise linear basis functions that are defined by a triangulation over the spatial area of interest \citep{lindgren2011}.
Commonly, this is referred to as a ``mesh''.
The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy and computational efficiency of the SPDE approach \citep{lindgren2011}.
Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an alternative parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
\[
\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right),
\]
where $\kappa$ is the Mat\'ern decorrelation rate as before.
We can calculate the marginal variance of the Mat\'ern random field as $\sigma^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.

@@

An additional delta-model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}.
The primary advantage is that both linear predictors use a log link so the linear predictors can be added in link space and the partial effect of a coefficient from both linear predictors can be combined.
In these models, the linear predictors represent a theoretical group number $n$ and a theoretical weight (e.g., mass) per group $w$ \citep{thorson2018poisson},
\[
\begin{aligned}
\log (n) &= \eta_1,\\
\log (w) &= \eta_2.
\end{aligned}
\]
Note that the first linear predictor has a log link, \emph{not} a logit link, and the linear predictor predicts group number $n$, \emph{not} positive observation probability $p$.
These get transformed \citep{thorson2018poisson} via
\[
\begin{aligned}
p &= 1 - \exp(- \exp( \log n)),\\
r &= \frac{n w}{p}.
\end{aligned}
\]
The first part is the complementary-log-log inverse link \citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but the group number $n$ also enters into the expected positive rate $r$.
These probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and positive observation likelihood as before
\[
\begin{aligned}
\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
\end{aligned}
\]

@@

<<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices'' or sometimes ``knots''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation to a point. A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly and supplying the object to \\code{make\\_mesh()}.", fig.width=4.5, fig.asp=1, out.width="3in">>=
mesh_pcod2 <- make_mesh(
  pcod,
  xy_cols = c("X", "Y"),
  fmesher_func = fmesher::fm_mesh_2d_inla,
  cutoff = 8,
  max.edge = c(10, 40),
  offset = c(10, 40)
)
plot(mesh_pcod2)
@

@@

\subsection{Exploring delta-model alternative families}

@@

On inspecting the model output with \fct{summary}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
We therefore consider a model where these are turned off in the second component of the delta model and choose this model.

@@

<<dog-depth-plot, echo=FALSE, fig.cap="The conditional effect of ocean bottom depth on Pacific Spiny Dogfish density. The line and shaded ribbon represent the mean and 95\\% confidence interval, respectively. Other fixed effects are held at constant values and the random fields are set to their expected value (zero).", out.width="4in">>=
ggplot(pred_depth, aes(
  exp(log_depth), exp(est),
  ymin = exp(est - 2 * est_se),
  ymax = exp(est + 2 * est_se))) +
  geom_ribbon(fill = "grey90") +
  geom_line() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(x = "Depth (m)", y = "Density")
@

@@

% \input{response}

>>>>

Data are often collected at particular locations in space or in space repeatedly over time. While such data are a rich source of information across many fields,
they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables \citep{cressie1993, diggle2007, cressie2011}.
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference \citep{legendre1989a, dormann2007}, can improve predictions \citep[e.g.,][]{shelton2014}, and can provide useful spatial summary statistics \citep[e.g.,][]{thorson2019d, barnett2021}.

@@

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \pkg{fmesher} or \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
\Rev{B2}Many individual features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date this full suite of useful features has not been integrated into a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and approximated in discrete time-intervals.

This paper describes the statistical models underlying \sdmTMB\ (Section~\ref{model-description}), explains how \sdmTMB\ is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \sdmTMB\ to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of links to other packages and future development directions (Section~\ref{discussion}).

@@

<<matern-range, warning=FALSE, message=FALSE, fig.width=9, out.width="\\textwidth", fig.asp=0.33, fig.align='center', fig.cap="Example Gaussian random fields for two range values. The range describes the distance at which spatial correlation decays to $\\approx 0.13$ in coordinate units (i.e., the distance at which two points are effectively independent). Panel (a) shows a shorter range than panel (b), which results in a ``wigglier'' surface. Panel (c) shows the Mat\\'ern function for these two range values. The dashed horizontal line shows a correlation of 0.13.">>=
predictor_dat <- expand.grid(
  x = seq(0, 1, length.out = 100),
  y = seq(0, 1, length.out = 100),
  year = seq_len(6)
)
sim_mesh <- make_mesh(predictor_dat, xy_cols = c("x", "y"), cutoff = 0.01)
s1 <- sdmTMB_simulate(
  formula = ~1,
  data = predictor_dat,
  mesh = sim_mesh,
  range = 0.2,
  phi = 0.1,
  sigma_O = 0.2,
  seed = 1,
  B = 0
)
sim_g1 <- ggplot(s1, aes(x, y, fill = mu)) +
  geom_raster(show.legend = FALSE) +
  scale_fill_viridis_c(option = "C") +
  coord_fixed(expand = FALSE) +
  theme_light() +
  # theme(axis.title = element_blank()) +
  ggtitle("(a) Range = 0.2") +
  labs(x = "X", y = "Y")

@@

Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}).
If we define $\lVert h \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\sigma_m^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\sigma_m^2$ is the marginal variance (magnitude of the random field ``wiggles''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
The parameter $\nu$ controls the smoothness of the covariance function.
In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}.
A more interpretable derived parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\mathrm{range} = \sqrt{8} / \kappa$), which has the empirically derived property of the distance at which correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).

@@

The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
\[
\Phi\left( s_j, s_k \right) =
  \frac{\sigma_m^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \bm{H} \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \bm{H} \right).
\]
Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{haskard2007, lindgren2011, thorson2015ices}.

@@

\Rev{B4b}For a user of \sdmTMB, the following are the important elements to understand.
First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
This means that GRF models can be computationally approximated as GMRFs.
By working with GMRFs, one can take advantage of theory developed to estimate their sparse precision matrix efficiently \citep{rue2005gmrf, lindgren2011} and avoid the inversion of large dense matrices.
Second, the SPDE involves piece-wise linear basis functions that are defined by a triangulation over the spatial area of interest \citep{lindgren2011}.
Commonly, this is referred to as a ``mesh''.
The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy and computational efficiency of the SPDE approach \citep{lindgren2011}.
Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an alternative parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
\[
\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right),
\]
where $\kappa$ is the Mat\'ern decorrelation rate as before.
We can calculate the marginal variance of the Mat\'ern random field as $\sigma_m^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.

@@

An additional delta model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}.
The primary advantage is that both linear predictors use a log link so the linear predictors can be added in link space and the partial effect of a coefficient from both linear predictors can be combined.
In these models, the linear predictors represent a theoretical group number $n$ and a theoretical weight (e.g., mass) per group $w$ \citep{thorson2018poisson},
\[
\begin{aligned}
\log (n) &= \eta_1,\\
\log (w) &= \eta_2.
\end{aligned}
\]
Note that the first linear predictor has a log link, \emph{not} a logit link, and the linear predictor predicts group number $n$, \emph{not} positive observation probability $p$.
These get transformed \citep{thorson2018poisson} via
\[
\begin{aligned}
p &= 1 - \exp(- \exp( \log n)),\\
r &= \frac{n w}{p}.
\end{aligned}
\]
The first part is the complementary-log-log inverse link \citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but the group number $n$ also enters into the expected positive rate $r$.
These probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and positive observation likelihood as before
\[
\begin{aligned}
\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
\end{aligned}
\]

@@

<<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices'' or ``knots''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation, to a point. A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly and supplying the object to \\code{make\\_mesh()}.", fig.width=4.5, fig.asp=1, out.width="3in">>=
mesh_pcod2 <- make_mesh(
  pcod,
  xy_cols = c("X", "Y"),
  fmesher_func = fmesher::fm_mesh_2d_inla,
  cutoff = 8,
  max.edge = c(10, 40),
  offset = c(10, 40)
)
plot(mesh_pcod2)
@

@@

\subsection{Exploring delta model alternative families}

@@

On inspecting the model output with \fct{summary}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
We therefore consider a model where these are turned off in the second component of the delta model and choose this model for further evaluation.

@@

<<dog-depth-plot, echo=FALSE, fig.cap="The conditional effect of ocean bottom depth on Pacific Spiny Dogfish population density. The line and shaded ribbon represent the mean and 95\\% confidence interval, respectively. Other fixed effects are held at constant values and the random fields are set to their expected value (zero).", out.width="4in">>=
ggplot(pred_depth, aes(
  exp(log_depth), exp(est),
  ymin = exp(est - 2 * est_se),
  ymax = exp(est + 2 * est_se))) +
  geom_ribbon(fill = "grey90") +
  geom_line() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(x = "Depth (m)", y = "Density")
@

@@

\input{response}
